Tutorial: Quantifying Spatially Resolved Hydration Thermodynamics Using Grid Inhomogeneous Solvation Theory [Article v\versionnumber]

### 2024-07-10 manuscript/manuscript.tex

\subsubsection{MD simulation}
\label{sec:md_simulation}
Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
Restraints should be used during the MD because IST theory is not well-defined for moving solute molecules.
We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NPT
&cntrl
	ntx=5, irest=1,
	ioutfm=1,
	ntb=2, iwrap=1,
	ntr=1, restraint_wt=100.0,
	restraintmask='!@H=&!:WAT',
	ntp=1, pres0=1.0, taup=1.0,
	ntc=2, ntf=2,
	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
	nstlim=50000000, dt=0.002,
	ntwr=50000, ntwx=5000, ntpr=5000,
/
\end{lstlisting}
This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
Note that these are very standard MD settings except for the restraints.
Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
-O \
-i 100ns-npt-restraint.in \
-o md-01.out \
-p solvated.parm7 \
-c EQUIL-DONE.rst \
-x md-01.nc \
-r md-01.ncrst \
-ref EQUIL-DONE.rst
\end{lstlisting}
\subsubsection{GIST analysis}

@@

\begin{lstlisting}[style=pymol]
load output/streptavidin/gist.pdb, streptavidin
as surface
color gray70, streptavidin
load output/streptavidin/gist-gO.dx, gO
isomesh gO_high, gO, 2
color slate, gO_high
\end{lstlisting}
A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent 
molecule, is found in Figure~\ref{fig-binding-pocket}.
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{figures/streptavidin_gO_high_surf.png}
	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density (i.e., bulk)}
	\label{fig-streptavidin_gO}
\end{figure}
\begin{figure}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
	\includegraphics[width=\linewidth]{figures/binding_pocket_E.png}
	\caption{Green: negative energy (based on $E_{sw} + 2E_{ww}$). Red: positive energy.}
	\label{fig-binding-pocket-energy}
    \end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\linewidth]{figures/binding_pocket_S.png}
		\caption{Blue: negative entropy}
		\label{fig-binding-pocket-entropy}
	\end{subfigure}
	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\square\angstrom}, except for 
	the positive energy, where they are divided by two to visually account for its overall lower density. Only voxels within \SI{4}{\angstrom} of biotin are shown.}
	\label{fig-binding-pocket}
\end{figure}

@@

	\begin{checklist}{Simulation settings}
		\textbf{Each number denotes the minimum setting, with the optimum in brackets.}
		\begin{itemize}
			\item Simulation time: \SIrange{10}{20}{\nano\second} (\SI{100}{\nano\second})
			\item Number of analyzed frames: \num{10000} (\num{100000})
			\item Restraints: typically \SI{100}{\kcalPerMolASqr} on solute heavy atoms.
		\end{itemize}
	\end{checklist}

>>>>

\subsubsection{MD simulation}
\label{sec:md_simulation}
Now, you can run a restrained MD simulation based on the equilibrated structure of biotin for the subsequent GIST analysis.
Restraints should be used during the MD because IST theory is not well-defined for moving solute molecules.
We apply a harmonic restraint of \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} to all heavy atoms to keep the system in place. Note that this is a rather strong restraint. While lower values are possible, make sure that the solute molecule is kept rigid during the simulation. As a minimum value, we recommend \SI{100}{\kilo\calorie\per\mole\per\angstrom\squared} restraints.
We run \SI{100}{\nano\second} of simulation and store the conformations every \SI{100}{\pico\second}.
A simulation length of at least \SI{20}{\nano\second} is recommended to obtain statistically independent snapshots of the water solvent molecules. 
An example Amber input file for GIST might look like this:
\begin{lstlisting}[style=amber-in]
restrained 100 ns NPT
&cntrl
	ntx=5, irest=1,
	ioutfm=1,
	ntb=2, iwrap=1,
	ntr=1, restraint_wt=100.0,
	restraintmask='!@H=&!:WAT',
	ntp=1, pres0=1.0, taup=1.0,
	ntc=2, ntf=2,
	ntt=3, tempi=300.0, temp0=300.0, gamma_ln=2,
	nstlim=50000000, dt=0.002,
	ntwr=50000, ntwx=5000, ntpr=5000,
/
\end{lstlisting}
This file will run a \SI{100}{\nano\second} NPT MD simulation while keeping all atoms except for hydrogens and waters restrained.
Note that these are very standard MD settings except for the restraints.
Run the MD using \software{pmemd.cuda}:
\begin{lstlisting}[style=bash]
pmemd.cuda \
-O \
-i 100ns-npt-restraint.in \
-o md-01.out \
-p solvated.parm7 \
-c EQUIL-DONE.rst \
-x md-01.nc \
-r md-01.ncrst \
-ref EQUIL-DONE.rst
\end{lstlisting}
\subsubsection{GIST analysis}

@@

\begin{lstlisting}[style=pymol]
load output/streptavidin/gist.pdb, streptavidin
as surface
color gray70, streptavidin
load output/streptavidin/gist-gO.dx, gO
isomesh gO_high, gO, 2
color slate, gO_high
\end{lstlisting}
A more sophisticated visualization, including the energy, entropy, and free energy of removing a single water solvent 
molecule, is found in Figure~\ref{fig-binding-pocket}.
\begin{figure}
	\centering
	\includegraphics[width=1\linewidth]{figures/streptavidin_gO_high_surf.png}
	\caption{Oxygen density around streptavidin at an isolevel of twice the reference density (i.e., bulk)}
	\label{fig-streptavidin_gO}
\end{figure}
\begin{figure}
	\centering
	\begin{subfigure}[b]{0.45\textwidth}
	\includegraphics[width=\linewidth]{figures/binding_pocket_E.png}
	\caption{GIST solvent energy (based on $E_{sw} + 2E_{ww}$).}
	\label{fig-binding-pocket-energy}
    \end{subfigure}
	\hfill
	\begin{subfigure}[b]{0.45\textwidth}
		\includegraphics[width=\linewidth]{figures/binding_pocket_S.png}
		\caption{GIST solvent entropy}
		\label{fig-binding-pocket-entropy}
	\end{subfigure}
	\caption{Biotin in the streptavidin binding pocket, with isosurfaces based on the GIST calculation on the holo structure. 
	Isolevels shown are ($\pm$) 0.25, 0.5, 1.0, 1.5, 2.0, 2.5, and 3.0 \si{\kilo\calorie\per\mole\per\cubic\angstrom}, except for 
	the positive energy, where they are divided by two to visually account for its overall lower density. The negative solvent energgy is shown in green, positive solvent energy in red and negative entropy in blue. Only voxels within \SI{4}{\angstrom} of biotin are shown.}
	\label{fig-binding-pocket}
\end{figure}

@@

	\begin{checklist}{Simulation settings}
		\textbf{Each number denotes the minimum setting, with the optimum in brackets.}
		\begin{itemize}
			\item Simulation time: \SIrange{10}{20}{\nano\second} (\SI{100}{\nano\second})
			\item Number of analyzed frames: \num{10000} (\num{100000})
			\item Restraints: typically \SI{100}{\kcalPerMolASqr} on solute heavy atoms. Minimum: \SI{10}{\kcalPerMolASqr}
		\end{itemize}
	\end{checklist}
